---
title: "Getting started with rtists"
vignette: >
  %\VignetteIndexEntry{Getting started with rtists}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
---

```{r}
#| label: setup
if (!requireNamespace("rtists", quietly = TRUE)) {
  knitr::opts_chunk$set(eval = FALSE)
} else {
  library(rtists)
  library(fiber)
}
```

## Overview

**rtists** (*R for Tissue Integrity Superimposed on Tractography Streamlines*)
provides interactive 3-D visualisation of tractography data defined in the
[fiber](https://github.com/tractoverse/fiber) package.  The central function is
`plot3d()`, a [S7](https://rconsortium.github.io/S7/) generic with methods for
the two core fiber classes:

| Class | Description |
|---|---|
| `fiber::streamline` | A single fibre tract — an ordered sequence of 3-D points plus optional metadata. |
| `fiber::bundle` | An ordered collection of streamlines representing a white-matter bundle. |

The interactive figures are produced by
[plotly](https://plotly.com/r/), so they can be panned, rotated, and zoomed
directly in the browser or the RStudio / Positron viewer pane.

---

## Data structures

A `streamline` stores three compartments, all accessible with the `@` operator:

- `@points` — an $n \times 3$ numeric matrix with columns `X`, `Y`, `Z`.
- `@point_data` — a named list of per-point numeric vectors (length $n$),
  e.g. fractional anisotropy (FA) sampled along the tract.
- `@streamline_data` — a named list of numeric scalars (length 1), e.g. a
  tract-level mean FA.

A `bundle` holds:

- `@streamlines` — a list of `streamline` objects.
- `@bundle_data` — a named list of bundle-level metadata.

### Creating a minimal example

```{r}
#| label: create-objects
pts <- matrix(
  c(
    0, 0, 0,
    1, 2, 1,
    2, 3, 3,
    3, 3, 5,
    4, 2, 6,
    5, 1, 7
  ),
  ncol = 3, byrow = TRUE,
  dimnames = list(NULL, c("X", "Y", "Z"))
)

sl <- streamline(
  points          = pts,
  point_data      = list(FA = c(0.25, 0.40, 0.65, 0.70, 0.55, 0.30)),
  streamline_data = list(mean_FA = 0.475)
)

sl
```

A bundle is just a list of streamlines wrapped by `bundle()`:

```{r}
#| label: create-bundle
# Simulate a small bundle of four slightly perturbed copies
set.seed(42)
streamlines <- lapply(seq_len(4), function(i) {
  noise <- matrix(rnorm(nrow(pts) * 3, sd = 0.15), ncol = 3)
  colnames(noise) <- c("X", "Y", "Z")
  streamline(
    points          = pts + noise,
    point_data      = list(FA = pmin(pmax(sl@point_data$FA + rnorm(6, sd = 0.05), 0), 1)),
    streamline_data = list(mean_FA = mean(sl@point_data$FA))
  )
})

bun <- bundle(streamlines)
bun
```

---

## Plotting with `plot3d()`

### Default: colour by local orientation

With no `color` argument, each point is coloured by the direction of the local
tangent vector.  Absolute values of the normalised $(dx, dy, dz)$ are mapped to
the R, G, B channels — the standard DTI convention (left–right = red,
anterior–posterior = green, superior–inferior = blue).

```{r}
#| label: fig-orientation
#| fig-cap: "Single streamline coloured by local fibre orientation."
plot3d(sl)
```

The same default works for a whole bundle:

```{r}
#| label: fig-bundle-orientation
#| fig-cap: "Bundle of four streamlines coloured by orientation."
plot3d(bun)
```

### Colour by per-point metadata

Pass any key from `@point_data` as the `color` argument.  Numeric values are
mapped to a continuous colour scale (default: `"Viridis"`).

```{r}
#| label: fig-point-fa
#| fig-cap: "Streamline coloured by per-point FA."
plot3d(sl, color = "FA")
```

### Colour by per-streamline metadata

Keys from `@streamline_data` are broadcast to every point of that streamline, 
so they work equally well as a colour aesthetic.  For a bundle this gives each
streamline a single uniform colour that reflects its tract-level scalar.

```{r}
#| label: fig-bundle-mean-fa
#| fig-cap: "Bundle coloured by per-streamline mean FA."
plot3d(bun, color = "mean_FA", palette = "RdYlBu")
```

### Fixed colour

Pass any CSS colour name or hex code to colour all lines identically.

```{r}
#| label: fig-fixed
#| fig-cap: "Bundle in a single fixed colour with reduced opacity."
plot3d(bun, color = "steelblue", opacity = 0.6)
```

---

## Appearance options

| Argument | Default | Description |
|---|---|---|
| `color` | `"orientation"` | Colour mode: `"orientation"`, metadata key, or a CSS/hex string. |
| `palette` | `"Viridis"` | Plotly / ColorBrewer scale used for continuous numeric metadata. |
| `linewidth` | `2` | Line width in pixels. |
| `opacity` | `0.5` | Global line opacity (0–1). |
| `...` | | Passed to `plotly::layout()`, e.g. `title = "Left CST"`. |

Extra `plotly::layout()` arguments can be supplied directly:

```{r}
#| label: fig-layout
#| fig-cap: "Custom title and increased line width."
plot3d(bun, color = "FA", linewidth = 4, title = "Example bundle — FA")
```
